---
title: "Calculate Activity Scores for all MSigDB-hallmark genesets"
author: "Birgitta"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  html_document:
  toc: true
highlight: haddock
editor_options: 
  chunk_output_type: console
---
# background
Calculate activity scores for all BRCA patients based on our curated TPM-mRNA expression data.

The Hallmark geneset from the Human MSigDB database was used. Version details:  v2022.1.Hs updated August 2022, Downloaded October, 5th 2022.

# libraries, working directories and functions
```{r libs_wd_fcts, message=FALSE, warning=FALSE}

# working directories
conf <- list()
conf$database_die <- file.path(~, "database")
conf$project_dir <- file.path(~, "activation_scores")

# libraries
library(plyr)
library(dplyr)
library(data.table)
library(matrixStats)
library(stringr)
library(openxlsx)
library(tibble)

```


# mRNA expression of geneset genes
In this section, expression values for the TCGA patients will be merged to the respective genes of each geneset. Resulting tables will be exported as .csv files eventually. 

```{r}
# Input: 
# read TCGA mRNA and confounder table from Supplementary tables of the manuscript. Only use those patients for further analysis, which are contained in the confounder table. 

mRNA <- join(confounder[,c("samp.ids", "subtype_mRNA", "sample_type")], mRNAtpm, by = "samp.ids", type = "inner")

## read genesets 
# file to be downloaded from the MSigDB database
genesets <- fread(file.path(conf$project_dir, "h.all.v2022.1.Hs.symbols.gmt"), sep = "\t", fill = TRUE, header= FALSE, data.table = FALSE) # read gmt file containing gene sets of interest here (gmt files can be downloaded from MSigDB. It's ok, to copy different lines in one file and rename it. )
genesets$V2 <- NULL
genesets <- as.data.frame(t(as.matrix(makeRn(genesets))))
names(genesets) <- str_sub(names(genesets), start = 10) # the "Hallmarks" pre-fix is cut of here from all geneset names to keep names a bit shorter. Optional step. 

#' ## merge Hallmark files with mRNA expression data
## merge Hallmark files with mRNA expression data -----
dflist <- apply(genesets, 2, function(x){
  df <- mRNA[,(names(mRNA) %in% x | names(mRNA) %in% c("samp.ids","subtype_mRNA","sample_type",	"ids"))]
  # df <- df%>% dplyr::select(c("ID", ), everything())
  return(df)
})

#check if merging was successful
begindf(dflist[[1]])
beginlist(dflist)
dflist[[1]][1:5,1:8]
ncol(dflist[[1]])

## export results as individual tables (needs to be performed, just as comment here to prevent problems while knitting the document)
# setwd(file.path(conf$project_dir, "Expression_GenesetGenes"))
# lapply(names(dflist), function(i){
#   write.csv2(dflist[[i]], sprintf("%s.csv", paste0(i, "_expr")), row.names = FALSE)
#   })

```


# Calculate activity score
In this sections, the files containing the expression for each gene of the respective geneset are read in again and used to calculate the actual activity score. 

```{r}
setwd(file.path(conf$project_dir, "Expression_GenesetGenes"))
files_exprs <- dir(pattern="_expr\\.csv$")

#### BRCA subtypes
subgroups <- c("all","Normal", "Tumour", "Basal", "LumA", "LumB", "Her2", "Normal_like")

## apply to all GSEA datasets (all files which are in folder "MSigDBListen")
act_score <- lapply(seq_along(files_exprs), function(i){
  # print(files_exprs[i])
  exprs <- read.csv2(files_exprs[i], stringsAsFactors = FALSE)
  
  ## apply to different BRCA subtypes, calculates subtype activity scores which strongly reflect the number of patients within a subtype
  subtype_score <-lapply(seq_along(subgroups), function(k){
    # print(subgroups[k])
    if(subgroups[k] == "all"){
      exprs_sub <- exprs
      } else if (subgroups[k] == "Tumour"){
        exprs_sub <- subset(exprs, exprs$subtype_mRNA  != "Normal")
        } else {
          exprs_sub <- subset(exprs ,exprs$subtype_mRNA  == subgroups[k])
        }
    exprs_sub[,2:4] <- NULL # remove useless annotation columns
    exprs_sub <- makeRn(exprs_sub) # make rownames from sample ids
    # exclude genes which have not been measured in any patient
    exprs_sub <- exprs_sub[,colSums(exprs_sub)>0]
    # use z-scores to calculate activation scores
    exprs_sub_calc <- apply(exprs_sub, 2, function(x){scale(x, center = TRUE, scale = TRUE)})
    rownames(exprs_sub_calc) <- rownames(exprs_sub)
    # use ranks to calculate activation scores
    # exprs_sub <- apply(exprs_sub, 2, rank)  # calculate rank for each patient for each gene -> each genes has values between 1 and number of patients
    exprs_sub <- as.data.frame(exprs_sub_calc)
    exprs_sub$gene_activity_score <- rowMedians(as.matrix(exprs_sub)) # calculate median rank for each patient --> = activity score
    # exprs_sub <- exprs_sub %>% select(c("gene_activity_score"), everything())
    exprs_sub <- make1col(exprs_sub)
    scores <- exprs_sub[,c("ids", "gene_activity_score")] #extract score information
    names(scores)[1] <- "samp.ids"
    # scores$ids <- gsub("\\.", "-", scores$ids)
    scores <- join(exprs[,c("samp.ids", "subtype_mRNA")], scores,  type = "right", by = "samp.ids")
     
    return(makeRn(scores))
    
    }) # close lapply for subtypes (all patients, basal,...)
  headlist(subtype_score)
  names(subtype_score) <- subgroups
  return(subtype_score)
  }) # close apply for list of Genesets

# str(act_score)
names(act_score) <- str_sub(files_exprs, end = -10) # the _expr.csv ending of the files is cut of here for meaningful naming of the activation scores

## generate summary tables for all and tumour patients 
### this section can be copied and adapted for the other subtypes, in case one of those is of interest. 

### combine & export activation score for all samples
all_pat <- lapply(seq_along(files_exprs), function(i){
  subgroup <- do.call(cbind, lapply(act_score[i], `[[`, "all")) # filter for dataframes named "all" in nested list --> exchange to subtype of interest if needed. 
    if (i < length(files_exprs)){
      subgroup <- subgroup[,2, drop = FALSE]
    } else if (i== length(files_exprs)){
      names(subgroup)[1] <- str_sub(names(subgroup)[1], start = 23) # cut off the dataset name in front of the subtype_mRNA
    }
  return(subgroup)
})
all_pat <- do.call("cbind", all_pat)
all_pat <- all_pat %>%
  select(c("subtype_mRNA"), everything()) %>%
  rownames_to_column("samp.ids")

# export activity score table for all patients (only as comment for knitting, do perform this step otherwise)
# fwrite(all_pat,file = file.path(conf$project_dir, "ActivationScores_AllSamp.csv"), sep = ";", dec = ",", row.names = FALSE)

# combine & export activation score for tumour samples
tumour_pat <- lapply(seq_along(files_exprs), function(i){
  subgroup <- do.call(cbind, lapply(act_score[i], `[[`, "Tumour")) # filter for dataframes named "Tumour" in nested list
    if (i < length(files_exprs)){
      subgroup <- subgroup[,2, drop = FALSE]
    } else if (i== length(files_exprs)){
      names(subgroup)[1] <- "subtype_mRNA"
    }
  return(subgroup)
})
tumour_pat <- do.call("cbind", tumour_pat)
tumour_pat <- tumour_pat %>%
  select(c("subtype_mRNA"), everything()) %>%
  rownames_to_column("samp.ids")

# fwrite(tumour_pat,file = file.path(conf$project_dir, "ActivitionScores_TumourSamp.csv"), sep = ";", dec = ",", row.names = FALSE)

```


# Session Info
```{r}
sessionInfo()
```
